---
title: "DMR Analysis - MDD (Major Depressive Disorder)"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

Differential m6A methylation analysis using pre-computed DMR results.

```{r setup, include=FALSE}
peak_counts_file <- snakemake@input$peak_counts
samplesheet_f    <- snakemake@input$samplesheet

library(data.table)
library(ggplot2)
library(dplyr)
library(tibble)
library(stringr)
library(pheatmap)
library(RColorBrewer)
library(reshape2)
```

# Load DMR data
```{r load-data}
# Load the pre-computed DMR table from TRESS
dmr_full <- read.table(gzfile(peak_counts_file), header=TRUE,
                       row.names=NULL, sep="\t", check.names=FALSE)

cat("DMR table dimensions:", dim(dmr_full), "\n")
cat("Columns:", colnames(dmr_full), "\n")

# The file contains paired IP+input columns per sample
# plus pre-computed stats: pvals, p.adj, rScore, lg.fc
head(dmr_full[, c("chr","start","end","lg.fc","pvals","p.adj")])
```
```{r load-samplesheet}
samplesheet <- fread(samplesheet_f)

# Get sample info from column names
ip_cols    <- colnames(dmr_full)[grepl("_m6A",   colnames(dmr_full))]
input_cols <- colnames(dmr_full)[grepl("_input", colnames(dmr_full))]

cat("IP samples:", length(ip_cols), "\n")
cat("Input samples:", length(input_cols), "\n")
```

# Compute DMRs per sex using log fold change

The file has overall stats. We compute sex-specific DMRs by
comparing IP/input ratios between MDD and control per sex.
```{r compute-sex-dmr}
compute_sex_dmr <- function(sex_group, dmr_full, samplesheet) {
  ss <- samplesheet[samplesheet$sex == sex_group, ]
  
  ctrl_samples <- ss$sample[ss$condition == "control"]
  mdd_samples  <- ss$sample[ss$condition == "MDD"]
  
  # Get IP columns for each group
  ctrl_ip <- paste0(ctrl_samples, "_m6A")
  ctrl_ip <- gsub("ctrl_1_m6A", "ctrl1_m6A", ctrl_ip)
  mdd_ip  <- paste0(mdd_samples,  "_m6A")
  ctrl_in <- paste0(ctrl_samples, "_input")
  mdd_in  <- paste0(mdd_samples,  "_input")
  
  # Fix any inconsistent naming (ctrl1 vs ctrl_1)
  ctrl_ip <- ctrl_ip[ctrl_ip %in% colnames(dmr_full)]
  ctrl_in <- ctrl_in[ctrl_in %in% colnames(dmr_full)]
  mdd_ip  <- mdd_ip[mdd_ip   %in% colnames(dmr_full)]
  mdd_in  <- mdd_in[mdd_in   %in% colnames(dmr_full)]
  # Keep only samples that have both IP and input
  ctrl_samples_ok <- intersect(gsub("_m6A","",ctrl_ip), gsub("_input","",ctrl_in))
  mdd_samples_ok  <- intersect(gsub("_m6A","",mdd_ip),  gsub("_input","",mdd_in))
  ctrl_ip <- paste0(ctrl_samples_ok, "_m6A")
  ctrl_in <- paste0(ctrl_samples_ok, "_input")
  mdd_ip  <- paste0(mdd_samples_ok,  "_m6A")
  mdd_in  <- paste0(mdd_samples_ok,  "_input")
  mdd_ip  <- mdd_ip[mdd_ip   %in% colnames(dmr_full)]
  ctrl_in <- ctrl_in[ctrl_in %in% colnames(dmr_full)]
  mdd_in  <- mdd_in[mdd_in   %in% colnames(dmr_full)]
  
  cat(paste0("\n--- ", sex_group, " ---\n"))
  cat("Control IP cols:", length(ctrl_ip), "\n")
  cat("MDD IP cols:", length(mdd_ip), "\n")
  
  if (length(ctrl_ip) == 0 | length(mdd_ip) == 0) {
    cat("WARNING: No matching columns found!\n")
    cat("Available m6A cols:", head(colnames(dmr_full)[grepl("m6A", colnames(dmr_full))]), "\n")
    return(NULL)
  }
  
  # Compute mean IP/input ratio per group
  ctrl_ratio <- rowMeans(log2(as.matrix(dmr_full[, ctrl_ip, drop=FALSE]) + 1) -
                         log2(as.matrix(dmr_full[, ctrl_in, drop=FALSE]) + 1))
  mdd_ratio  <- rowMeans(log2(as.matrix(dmr_full[, mdd_ip,  drop=FALSE]) + 1) -
                         log2(as.matrix(dmr_full[, mdd_in,  drop=FALSE]) + 1))
  
  # Log fold change: MDD vs control
  lfc <- mdd_ratio - ctrl_ratio
  
  # Use overall p-values from TRESS (already computed)
  result <- data.frame(
    peak   = paste0(dmr_full$chr, ":", dmr_full$start, "-", dmr_full$end),
    chr    = dmr_full$chr,
    start  = dmr_full$start,
    end    = dmr_full$end,
    log2FoldChange = lfc,
    pvalue = dmr_full$pvals,
    padj   = dmr_full$p.adj,
    sex    = sex_group
  )
  
  # Significant DMRs: nominal p < 0.05, FC > 1.2
  sig <- result %>%
    filter(!is.na(pvalue), pvalue < 0.05, abs(log2FoldChange) > log2(1.5))
  
  cat(paste0("Total DMRs tested: ", nrow(result), "\n"))
  cat(paste0("Significant DMRs: ", nrow(sig), "\n"))
  cat(paste0("  Hypermethylated: ", sum(sig$log2FoldChange > 0), "\n"))
  cat(paste0("  Hypomethylated:  ", sum(sig$log2FoldChange < 0), "\n"))
  
  list(all = result, sig = sig)
}

dmr_male   <- compute_sex_dmr("male",   dmr_full, samplesheet)
dmr_female <- compute_sex_dmr("female", dmr_full, samplesheet)
```

# Figure 2A — Volcano plots (Male blue, Female pink)

Replicates Figure 2A from Mitsuhashi et al. 2024.
```{r fig2a, fig.width=14, fig.height=6}
# Prepare plot data
v_male <- dmr_male$all %>%
  mutate(
    sex = "Male MDD",
    sig = pvalue < 0.05 & abs(log2FoldChange) > log2(1.2)
  )

v_female <- dmr_female$all %>%
  mutate(
    sex = "Female MDD",
    sig = pvalue < 0.05 & abs(log2FoldChange) > log2(1.2)
  )

v_both <- bind_rows(v_male, v_female) %>%
  mutate(sex = factor(sex, levels = c("Male MDD", "Female MDD")))

# Paper colors: blue for male, pink for female
ggplot(v_both, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(data = filter(v_both, !sig),
             color = "grey80", size = 0.5, alpha = 0.4) +
  geom_point(data = filter(v_both, sig & sex == "Male MDD"),
             color = "#4393C3", size = 1.5, alpha = 0.8) +
  geom_point(data = filter(v_both, sig & sex == "Female MDD"),
             color = "#F4A582", size = 1.5, alpha = 0.8) +
  facet_wrap(~ sex) +
  geom_vline(xintercept = c(-log2(1.2), log2(1.2)),
             linetype = "dashed", color = "black", linewidth = 0.3) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "black", linewidth = 0.3) +
  xlim(-3, 3) +
  theme_bw() +
  theme(
    strip.text   = element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill = "white"),
    legend.position = "none",
    plot.title   = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10)
  ) +
  labs(
    title    = "(A) Volcano plot showing DMRs in male and female MDD",
    subtitle = paste0(
      "Male MDD: ", nrow(dmr_male$sig), " DMRs (blue)  |  ",
      "Female MDD: ", nrow(dmr_female$sig), " DMRs (pink)"
    ),
    x = "Log2 Fold Change (MDD / control)",
    y = "-log10(p-value)"
  )
```

# DMR overlap: unique to male vs female
```{r overlap}
male_peaks   <- dmr_male$sig$peak
female_peaks <- dmr_female$sig$peak
shared       <- intersect(male_peaks, female_peaks)
only_male    <- setdiff(male_peaks, female_peaks)
only_female  <- setdiff(female_peaks, male_peaks)

cat(paste0("Male-only DMRs:   ", length(only_male), "\n"))
cat(paste0("Female-only DMRs: ", length(only_female), "\n"))
cat(paste0("Shared DMRs:      ", length(shared), "\n"))
```

# Summary bar chart
```{r barplot, fig.width=7, fig.height=5}
dmr_summary <- data.frame(
  Sex       = rep(c("Male MDD", "Female MDD"), each = 2),
  Direction = rep(c("Hypermethylated", "Hypomethylated"), 2),
  Count     = c(
    sum(dmr_male$sig$log2FoldChange   > 0),
    sum(dmr_male$sig$log2FoldChange   < 0),
    sum(dmr_female$sig$log2FoldChange > 0),
    sum(dmr_female$sig$log2FoldChange < 0)
  )
)

ggplot(dmr_summary, aes(x = Sex, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c(
    "Hypermethylated" = "#D6604D",
    "Hypomethylated"  = "#4393C3"
  )) +
  theme_bw() +
  labs(y = "Number of DMRs", x = "") +
  theme(legend.position = "none")
```

# Heatmap of top DMRs
```{r heatmap, fig.width=12, fig.height=7}
# Get top DMR peaks
top_peaks <- union(
  head(dmr_male$sig   %>% arrange(pvalue) %>% pull(peak), 30),
  head(dmr_female$sig %>% arrange(pvalue) %>% pull(peak), 30)
)

# Get IP counts for these peaks
all_ip_cols <- colnames(dmr_full)[grepl("_m6A", colnames(dmr_full))]
peak_ids    <- paste0(dmr_full$chr, ":", dmr_full$start, "-", dmr_full$end)

idx <- which(peak_ids %in% top_peaks)

if (length(idx) > 0) {
  mat <- log2(as.matrix(dmr_full[idx, all_ip_cols]) + 1)
  rownames(mat) <- peak_ids[idx]
  
  # Clean colnames
  colnames(mat) <- gsub("_m6A", "", colnames(mat))
  
  # Annotation
  ann_col <- samplesheet %>%
    filter(sample %in% colnames(mat)) %>%
    dplyr::select(sample, sex, condition) %>%
    column_to_rownames("sample")
  mat <- mat[, rownames(ann_col), drop=FALSE]
  
  pheatmap::pheatmap(mat,
    scale = "row", show_rownames = FALSE,
    annotation_col = ann_col,
    annotation_colors = list(
      condition = c(control = "#2166AC", MDD = "#D6604D"),
      sex       = c(male = "#4393C3", female = "#F4A582")
    ),
    main = "Top DMRs - m6A IP counts (log2)",
    cluster_cols = TRUE, fontsize_col = 6
  )
}
```

# GO enrichment of DMR genes
```{r gsea, warning=FALSE, message=FALSE}
library(fgsea)
library(msigdbr)

msig_human <- msigdbr::msigdbr(species = "Homo sapiens", category = "C5") %>%
  dplyr::select(gs_name, gene_symbol) %>% dplyr::distinct()
pathways <- split(msig_human$gene_symbol, msig_human$gs_name)

run_gsea <- function(dmr_result, label) {
  ranks <- setNames(
    -log10(dmr_result$all$pvalue + 1e-10) * sign(dmr_result$all$log2FoldChange),
    dmr_result$all$peak
  )
  ranks <- sort(ranks[!is.na(ranks)], decreasing = TRUE)
  
  set.seed(1)
  fg <- fgsea(pathways = pathways, stats = ranks,
              minSize = 15, maxSize = 500, eps = 0) %>%
    as_tibble() %>% arrange(pval)
  
  cat(paste0("\nTop pathways - ", label, ":\n"))
  print(head(fg[, c("pathway","NES","pval","padj")], 10))
  fg
}

fg_male   <- run_gsea(dmr_male,   "Male DMR")
fg_female <- run_gsea(dmr_female, "Female DMR")
```

## Dotplot male DMRs
```{r dotplot-male, fig.width=14, fig.height=7}
plot_dotplot <- function(fg, title) {
  top_p <- bind_rows(
    fg %>% filter(NES > 0) %>% arrange(pval) %>% head(15),
    fg %>% filter(NES < 0) %>% arrange(pval) %>% head(15)
  ) %>% arrange(NES)
  
  top_p$short <- str_trunc(gsub("GOBP_|GOCC_|GOMF_","",top_p$pathway), 55)
  
  ggplot(top_p, aes(x=NES, y=reorder(short,NES),
                    size=-log10(pval), color=NES>0)) +
    geom_point() +
    scale_color_manual(values=c("TRUE"="#D6604D","FALSE"="#4393C3"),
      labels=c("TRUE"="Hypermethylated","FALSE"="Hypomethylated")) +
    geom_vline(xintercept=0, linetype="dashed", color="grey50") +
    theme_minimal() +
    theme(axis.text.y=element_text(size=7), legend.title=element_blank()) +
    labs(x="NES", y="GO Pathway", title=title, size="-log10(p)")
}

plot_dotplot(fg_male, "DMR pathway enrichment - Male MDD")
```

## Dotplot female DMRs
```{r dotplot-female, fig.width=14, fig.height=7}
plot_dotplot(fg_female, "DMR pathway enrichment - Female MDD")
```
```{r session-info}
sessionInfo()
```