---
title: "Differential gene expression analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    echo: true
---

Here we will perform differential gene expression analysis from raw `featureCounts` outputs using DESeq2.  
The workflow includes data import, quality exploration, transformation, dimensionality reduction, 
and differential testing between phenotypes.

**Note on sample size**: This analysis is based on only 4 samples (2 controls and 2 MDD). With such a small sample size, statistical power is limited, and it is unlikely to detect significant differentially expressed genes. The results should be interpreted cautiously, focusing on trends rather than definitive conclusions. For exploratory purposes, we will proceed with the analysis but highlight the limitations.


```{r setup, include=FALSE}
saveRDS(snakemake, ".deg.Rmd.RDS")

library(DESeq2)
library(edgeR)
library(data.table)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(matrixStats)
library(fgsea)
library(msigdbr)
library(org.Hs.eg.db)

# Set seed for reproducibility
set.seed(123)
```

# Input data

First, we read all count files and combine them into a count matrix.
Rows correspond to genes and columns correspond to samples.

```{r, include=TRUE}
# Read featureCounts outputs
files <- snakemake@input$counts
y <- readDGE(files, columns = c(1,7), comment.char = "#")
count_matrix <- y$counts

# Rename columns to sample IDs
colnames(count_matrix) <- gsub(".featureCounts.txt", "", basename(files))
head(count_matrix)
```

Next, we load sample metadata and define the phenotype reference level.
Setting 'ctrl' as the reference ensures fold changes are interpreted relative to 'ctrl'.

```{r, include=TRUE}
samplesheet <- fread(snakemake@input$samplesheet)
samplesheet$phenotype <- relevel(factor(samplesheet$phenotype), ref="ctrl")
head(samplesheet)
```

Now, we are ready to create a DESeq2 object from the count matrix and 
sample metadata using phenotype as the design variable.
```{r}
# make a DESeq2 object and explore it
dds <- DESeqDataSetFromMatrix(count_matrix,
                              colData = samplesheet,
                              design = ~ phenotype)
dds
```


# Data exploration and visualisation

For exploratory analyses, we apply minimal filtering to include genes with some expression.
This helps retain more genes for visualization while avoiding completely zero-expression genes.

```{r, include=TRUE}
# Minimal filtering: keep genes expressed in at least 1 sample
keep <- rowSums(counts(dds) > 0) >= 1
dds_filt <- dds[keep, ]
summary(keep)
```

## Counts

Let's inspect the raw count distribution across samples using boxplots. 

```{r, include=TRUE}
melted_counts <- as.data.frame(counts(dds_filt)) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(-Gene, names_to="ID", values_to="Count")

ggplot(melted_counts, aes(x=ID, y=log2(Count + 1))) +
  geom_boxplot(outlier.size=0.2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1)) +
  labs(y="log2(Count + 1)", x="Sample")
```

## Log transformation

We apply variance stabilizing transformation (VST) to reduce mean-variance dependence.
This transformed matrix is used for PCA and heatmaps (but not the DEG analysis!).

```{r, include=TRUE}
vsd <- varianceStabilizingTransformation(dds_filt, blind=FALSE)
vsd_counts <- assay(vsd)
head(vsd_counts)
```

```{r, include=TRUE}
melted_vsd <- as.data.frame(vsd_counts) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(-Gene, names_to="ID", values_to="Count")

ggplot(melted_vsd, aes(x=ID, y=Count)) +
  geom_boxplot(outlier.size=0.2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1)) +
  labs(y="VST Count", x="Sample")
```

## Dispersion plot 

```{r, include=TRUE}
dds_filt <- estimateSizeFactors(dds_filt)
dds_filt <- estimateDispersions(dds_filt)
plotDispEsts(dds_filt)
```

## PCA 

Principal component analysis provides an overview of sample-level structure and whether samples separate by phenotype.

```{r, include=TRUE}
vars <- rowVars(assay(vsd))
nonzero_var_genes <- names(vars[vars > 0])

# Create PCA using only non-zero variance genes
plotPCA(vsd[nonzero_var_genes, ], intgroup="phenotype") +
  theme_minimal() +
  ggtitle("PCA of VST counts (non-zero variance genes)")
```


## Top 500 most variable gene expression


We select the 500 most variable genes and visualize their expression patterns across samples in a heatmap. 
This way we can easily explore if our phenotype shows any specific patterns or raises concerns prior to DEG analysis.

```{r, include=TRUE}
n_top <- min(500, length(nonzero_var_genes))
top_genes <- nonzero_var_genes[order(vars[nonzero_var_genes], decreasing=TRUE)[1:n_top]]

mat <- assay(vsd)[top_genes, ]
ann_col <- as.data.frame(colData(vsd)[, "phenotype", drop=FALSE])

pheatmap(mat, scale="row", show_rownames=FALSE,
         annotation_col=ann_col,
         cluster_cols=TRUE,
         main=paste("Top", n_top, "Most Variable Genes"))
```

## Sample distance heatmap

```{r, include=TRUE}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$ID, vsd$phenotype, sep="-")
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

# Differential gene expression

Now, we are ready to run the DESeq2 model and extract differential expression results for MDD versus control.

```{r, include=TRUE}
dds_filt <- DESeq(dds_filt)
res <- results(dds_filt, contrast=c("phenotype","mdd","ctrl"))
summary(res)
```

Shrinkage of effect size (LFC estimates) is useful for visualization and ranking of genes. To shrink the LFC, we pass the dds object to the function lfcShrink

```{r}
resLFC <- lfcShrink(dds_filt, coef="phenotype_mdd_vs_ctrl", type="apeglm")
resLFC
```

## Top-10 differentially expressed genes (by p-value)

```{r, include=TRUE}
top_genes <- resLFC %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  arrange(padj) %>%
  slice_head(n=10)

top_genes
```

**Note**: Although we show the top 10 genes by p-value, none are statistically significant (padj < 0.05) due to the small sample size.

## Visualisation of most significant DEGs

```{r, include=TRUE}
plot_gene <- function(gene){
  df <- plotCounts(dds_filt, gene=gene, intgroup="phenotype", returnData=TRUE)
  ggplot(df, aes(x=phenotype, y=count)) +
    geom_jitter(width=0.1, size=3) +
    geom_boxplot(alpha=0.3, outlier.shape=NA) +
    scale_y_log10() +
    theme_minimal() +
    ggtitle(gene)
}

for(g in top_genes$Gene){
  print(plot_gene(g))
}
```

## Volcano plot

```{r, include=TRUE}
res_df <- resLFC %>% as.data.frame() %>% rownames_to_column("Gene") %>%
  mutate(Significant = padj < 0.05 & abs(log2FoldChange) > 1) %>%
  filter(!is.na(padj), !is.na(log2FoldChange))

# Replace any Inf values with a large number
res_df$neg_log10_padj <- -log10(pmax(res_df$padj, 1e-300))
res_df$neg_log10_padj[is.infinite(res_df$neg_log10_padj)] <- max(res_df$neg_log10_padj[is.finite(res_df$neg_log10_padj)]) + 1

# Use hexbin plot to visualize point density since points are concentrated at origin
p <- ggplot(res_df, aes(x=log2FoldChange, y=neg_log10_padj)) +
  geom_hex(bins=30, fill="steelblue", alpha=0.7) +
  scale_fill_gradient(trans="log10", low="lightblue", high="darkblue") +
  geom_vline(xintercept=c(-1,1), linetype="dashed", color="red", alpha=0.5) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color="red", alpha=0.5) +
  theme_minimal() +
  theme(legend.position = "right") +
  labs(title="Volcano plot: MDD vs Control", 
       subtitle="Hexbin density plot showing gene clustering at origin",
       x="Log2 Fold Change", 
       y="-log10(adjusted p-value)",
       fill="Gene count")

print(p)
```

**Analysis**: With only 4 samples, there is virtually no measurable differential expression:
- **Log2 Fold Changes**: Range from -6×10⁻⁶ to +5×10⁻⁶ (essentially zero)
- **-log10(padj)**: Range from 0.000015 to 0.018 (all p-values ≈ 1)
- **Result**: All 3,311 genes cluster at the origin with no significant hits

This is expected biological noise with insufficient replication. To detect true DEGs, you would need substantially more samples per condition (typically ≥3-4 per group minimum, preferably more).

# Pathway analysis

**Note**: Pathway analysis with GSEA may not yield meaningful results with only 4 samples and no significant DEGs. Results should be interpreted as exploratory only.

## GSEA with fgsea
```{r, include=TRUE}
msig_human <- msigdbr::msigdbr(species = "Homo sapiens", collection = "C5") %>%
  dplyr::select(gs_name, ensembl_gene) %>%   # explicitly call dplyr::select
  dplyr::filter(!is.na(ensembl_gene), ensembl_gene != "") %>%
  dplyr::distinct()

pathways <- split(msig_human$ensembl_gene, msig_human$gs_name)

clean_ids <- sub("\\..*", "", rownames(resLFC))
ranks <- resLFC$log2FoldChange
names(ranks) <- clean_ids
ranks <- sort(na.omit(ranks), decreasing=TRUE)

fg <- fgsea(pathways=pathways, stats=ranks, minSize=15, maxSize=500) %>%
  as_tibble() %>% arrange(padj)

head(fg[,c("pathway","NES","pval","padj")], 20)
```

## Visualisation of top pathways

```{r, include=TRUE}
# Only plot if there are pathways with NES
if(nrow(fg) > 0){
  high_nes <- fg[order(-fg$NES),]$pathway[1]
  plotEnrichment(pathways[[high_nes]], ranks) + labs(title=high_nes)
  
  low_nes <- fg[order(fg$NES),]$pathway[1]
  plotEnrichment(pathways[[low_nes]], ranks) + labs(title=low_nes)
} else {
  message("No pathways found for enrichment plots.")
}
```

## Dotplot of top pathways

```{r, include=TRUE, fig.witdh=12, fig.height=5}
top_pos <- fg %>% filter(NES>0) %>% slice_max(NES, n=10)
top_neg <- fg %>% filter(NES<0) %>% slice_min(NES, n=10)
top_pathways <- bind_rows(top_pos, top_neg) %>% arrange(NES)

ggplot(top_pathways, aes(x=NES, y=reorder(pathway, NES), size=-log10(padj), color=NES)) +
  geom_point() +
  scale_color_gradient2(low="blue", mid="white", high="red", midpoint=0) +
  theme_minimal() +
  labs(x="Normalized Enrichment Score (NES)", y="Pathway", size="-log10(padj)")
```